Technical Dimensions of Programming Systems

### 2023-02-17 prog-2022/prog22-master.tex
Final camera-ready changes

% Class scrartcl Warning: seems someone has broken package `auxhook'.
% Usually this happens, if `auxhook' is loaded or used implicitly or explicitly
% by patching \document or via etoolbox command \AtEndPreamble. Trying an
% emergency workaround. You can avoid this warning adding:
\usepackage{auxhook}
% before \begin{document} on input line 6.
\usepackage{changepage} % For the single-page summary table PDF insert.
\usepackage{pdfpages}
\usepackage{pifont} % For checkmark / cross symbols for Appendix table :)
\usepackage{amsthm} % For numberless Definition
\usepackage{epigraph} % For the opening quotes

@@

\author{Joel Jakubovic}
\affiliation{
University of Kent, Canterbury, UK
\email{jdj9@kent.ac.uk}
}
\author{Jonathan Edwards}
\affiliation{\email{jonathanmedwards@gmail.com}}
\author{Tomas Petricek}
\affiliation{
University of Kent, Canterbury, UK \\
Charles University, Prague, Czechia
\email{tomas@tomasp.net}
}

@@

\begin{abstract}
  \emph{Context.} Programming requires much more than just writing code in a programming language. It is usually done in the context of a stateful environment, by interacting with a system through a graphical user interface. Yet, this wide space of possibilities lacks a common structure for navigation. Work on programming systems fails to form a coherent body of research, making it hard to improve on past work and advance the state of the art.

  \emph{Inquiry.} In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems;} we believe that programming systems deserve a theory too.

  \emph{Approach.} We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.

  \emph{Knowledge.} We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction. Technical dimensions capture characteristics that may be studied, compared and advanced independently. This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.

  \emph{Grounding.} Our framework is derived using a qualitative analysis of past programming systems. We outline two concrete ways of using our framework. First, we show how it can analyze a recently developed novel programming system. Then, we use it to identify an interesting unexplored point in the design space of programming systems.

  \emph{Importance.} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected. They are informal, guided by the personal vision of their authors and thus are only evaluable and comparable on the basis of individual experience using them. By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
\end{abstract}

@@

\newcommand{\joel}[1]{}
\newcommand{\note}[1]{}
\newcommand{\tp}[1]{}
\newcommand{\mybox}[1]{\noindent\fbox{\parbox{\textwidth}{#1}}}
\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
\newtheorem*{defn}{Definition}
%\newcommand{\hypertarget}[1]{}

>>>>

%\usepackage{changepage}
%\usepackage{lscape} % For the summary table
\usepackage{pdfpages}
\usepackage{pifont} % For checkmark / cross symbols for Appendix table :)
\usepackage{amsthm} % For numberless Definition
\usepackage{epigraph} % For the opening quotes

@@

\author[a]{Joel Jakubovic}
\authorinfo[joel]{is Tomas' final-year PhD student interested in overcoming pervasive ``premature commitment'' in programming. His forthcoming dissertation is about creating programming systems that combine self-sustainability and notational freedom. Send him an e-mail at \email{jdj9@kent.ac.uk} and read his blog at \href{https://programmingmadecomplicated.wordpress.com}{\texttt{programmingmadecomplicated.wordpress.com}}.}
\author[b]{Jonathan Edwards}
\authorinfo[jonathan]{is an independent researcher working on drastically simplifying programming. He is known for his Subtext series of programming language experiments and his blog at \href{https://alarmingdevelopment.org}{\texttt{alarmingdevelopment.org}}. He has been a researcher at MIT CSAIL and CDG/HARC. He tweets \texttt{@jonathoda} and can be reached at \email{jonathanmedwards@gmail.com}.}
\author[a,c]{Tomas Petricek}
\authorinfo[tomas]{is an assistant professor at Charles University. He is interested in finding easier and more accessible ways of thinking about programming. To do so, he combines technical work on programming systems and tools with research into history and philosophy of science. His work can be found at \href{https://tomasp.net}{\texttt{tomasp.net}} and he can be reached at \email{tomas@tomasp.net}.}
\affiliation[a]{University of Kent, Canterbury, UK}
\affiliation[b]{Independent}
\affiliation[c]{Charles University, Prague, Czechia}

@@

\newcommand{\ciakgi}[1]{}

\begin{abstract}
  \ciakgi{Context} Programming requires much more than just writing code in a programming language. It is usually done in the context of a stateful environment, by interacting with a system through a graphical user interface. Yet, this wide space of possibilities lacks a common structure for navigation. Work on programming systems fails to form a coherent body of research, making it hard to improve on past work and advance the state of the art.

  \ciakgi{Inquiry} In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems;} we believe that programming systems deserve a theory too.

  \ciakgi{Approach} We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.

  \ciakgi{Knowledge} We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction. Technical dimensions capture characteristics that may be studied, compared and advanced independently. This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.

  \ciakgi{Grounding} Our framework is derived using a qualitative analysis of past programming systems. We outline two concrete ways of using our framework. First, we show how it can analyze a recently developed novel programming system. Then, we use it to identify an interesting unexplored point in the design space of programming systems.

  \ciakgi{Importance} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected. They are informal, guided by the personal vision of their authors and thus are only evaluable and comparable on the basis of individual experience using them. By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
\end{abstract}

@@

\newcommand{\joel}[1]{}
\newcommand{\note}[1]{}
\newcommand{\tp}[1]{}
%\newcommand{\mybox}[1]{\noindent\fbox{\parbox{\textwidth}{#1}}}
\newcommand{\mybox}[1]{\textbf{#1}}
\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
\newtheorem*{defn}{Definition}
%\newcommand{\hypertarget}[1]{}

### 2023-02-17 prog-2022/summary-table-old.tex
Final camera-ready changes

\begin{table}
\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline CUSTOMIZABILITY &
\summ{Once a program exists in the system, how can it be extended and modified?}
Staging of Customization &
Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
Externalizability &
Which portions of the system's state can be referenced and transferred to/from it? &
None (state is private) vs. all state exposed as human-legible, CSS-like addressing \\
Additive Authoring &
How far can the system's behavior be changed by \emph{adding} expressions? &
None (requires power to change original) vs. full (anything can be overridden repeatedly) \\
Self-Sustainability &
How far can the system’s behavior be changed from within? &
None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
\hline COMPLEXITY &
\summ{How does the system structure complexity and what level of detail is required?}
Factoring of Complexity &
What programming details are hidden in reusable components and how? &
Domain-specific (more hiding) vs. general-purpose (less hiding) programming \\
Level of Automation &
What part of program logic does not need to be explicitly specified? &
Garbage collection (low-tech) vs. Prolog engine (hi-tech) \\
\hline ERRORS &
\summ{What does the system consider to be an error; how are they prevented and handled?}
Error Detection &
What errors can be detected in which feedback loops, and how? &
Human inspection in live coding vs. partial automation in static typing \\
Error Response &
How does the system respond when an error is detected? &
Does it stop, recover automatically, ignore the error or ask the user how to continue? \\
\hline ADOPTABILITY &
\summ{How does the system facilitate or obstruct adoption by both individuals and communities?}
Learnability &
What is the attitude towards the \emph{learning curve} and what is the target audience? &
HyperCard for the general public vs. FORTRAN for scientists \\
Sociability &
What are the social and economic factors that make the system the way it is? &
Funding, volunteers, code sharing, Q/A sites vs. documentation, sense of belonging \\
\hline
\end{tabular}
\caption{\label{summary-table2} Quick reference sheet for our set of Technical Dimensions.}
\end{table}

>>>>

\begin{table}
\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline CUSTOMIZABILITY &
\summ{Once a program exists in the system, how can it be extended and modified?}
Staging of Customization &
Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
Externalizability / Additive Authoring &
Which portions of the system's state can be referenced and transferred to/from it? How far can the system's behavior be changed by \emph{adding} expressions? &
State is: (\emph{i}) private and hidden; (\emph{ii}) externalizable and overwriteable; (\emph{iii}) externalized and \emph{overridable} by addition \\
Self-Sustainability &
How far can the system’s behavior be changed from within? &
None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
\hline COMPLEXITY &
\summ{How does the system structure complexity and what level of detail is required?}
Factoring of Complexity &
What programming details are hidden in reusable components and how? &
Domain-specific (more hiding) vs. general-purpose (less hiding) programming \\
Level of Automation &
What part of program logic does not need to be explicitly specified? &
Garbage collection (low-tech) vs. Prolog engine (hi-tech) \\
\hline ERRORS &
\summ{What does the system consider to be an error; how are they prevented and handled?}
Error Detection &
What errors can be detected in which feedback loops, and how? &
Human inspection in live coding vs. partial automation in static typing \\
Error Response &
How does the system respond when an error is detected? &
Does it stop, recover automatically, ignore the error or ask the user how to continue? \\
\hline ADOPTABILITY &
\summ{How does the system facilitate or obstruct adoption by both individuals and communities?}
Learnability &
What is the attitude towards the \emph{learning curve} and what is the target audience? &
HyperCard for the general public vs. FORTRAN for scientists \\
Sociability &
What are the social and economic factors that make the system the way it is? &
Funding, volunteers, code sharing, Q/A sites vs. documentation, sense of belonging \\
\hline
\end{tabular}
\caption{\label{summary-table2} Quick reference sheet for our set of Technical Dimensions.}
\end{table}

### 2023-02-17 prog-2022/tech-dims-table-portrait.tex
Final camera-ready changes

\newcommand{\summ}[1]{\multicolumn{2}{p{0.8\linewidth}}{#1} \\ \hline}
\setlength\extrarowheight{2pt}
\begin{table}
%\centering
\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline INTERACTION &
\summ{How do users execute their ideas, evaluate the result, and generate new ideas in response?}
Feedback Loops &
How wide are the various gulfs of \emph{execution} and \emph{evaluation} and how are they related? &
Immediate Feedback (short) vs. batch mode (long) gulf of evaluation \\
Modes of Interaction &
Which sets of feedback loops only occur together? &
Setup vs. editing vs. debugging \\
Abstraction Construction &
How do we go from abstractions to concrete examples and vice versa? &
Programming by Example vs. first principles \\
\hline NOTATION &
\summ{How are the different textual / visual programming notations related?}
Notational Structure &
What notations are used to program the system and how are they related? &
Notations overlap and need sync vs. complement each other \\
Surface / Internal Notations &
What is the connection between what a user sees and what a computer program sees? &
Sequence Editing vs. Rendering, Structure Editing vs. Recovery \\
Primary / Secondary Notations &
Is one notation more important than others? &
Secondary build scripts vs. visual editor and code on equal footing in Flash \\
Expression Geography &
Do similar expressions encode similar programs? &
Concise yet error-prone vs. explicit yet verbose \\
Uniformity of Notations &
Does the notation use a small or a large number of basic concepts? &
Lisp S-expressions vs. English-like textual notations \\
\hline ARCHITECTURE &
\summ{What is the system's attitude to its own organization and that of the outside world?}
Conceptual Outlook &
Does the system present as \emph{designed} or \emph{improvised}? &
Integrity (Everything is a X) vs. openness (compatible mixtures) \\
Composability &
What are the primitives? How can they be combined to achieve novel behaviors? &
Sequence, selection, repetition, function abstraction, recursion, logical connectives \\
Convenience &
Which wheels do users not need to reinvent? &
Small vs. expansive standard libraries \\
Commonality &
How much is \emph{common structure} explicitly marked as such? &
Common structure is redundantly flattened vs. factored out\\
\hline
\end{tabular}
\caption{\label{summary-table1} Quick reference sheet for our set of Technical Dimensions.}
\end{table}

\begin{table}
\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline CUSTOMIZABILITY &
\summ{Once a program exists in the system, how can it be extended and modified?}
Staging of Customization &
Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
Externalizability / Additive Authoring &
Which portions of the system's state can be referenced and transferred to/from it? How far can the system's behavior be changed by \emph{adding} expressions? &
State is: (\emph{i}) private and hidden; (\emph{ii}) externalizable and overwriteable; (\emph{iii}) externalized and \emph{overridable} by addition \\
Self-Sustainability &
How far can the system’s behavior be changed from within? &
None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
\hline COMPLEXITY &
\summ{How does the system structure complexity and what level of detail is required?}
Factoring of Complexity &
What programming details are hidden in reusable components and how? &
Domain-specific (more hiding) vs. general-purpose (less hiding) programming \\
Level of Automation &
What part of program logic does not need to be explicitly specified? &
Garbage collection (low-tech) vs. Prolog engine (hi-tech) \\
\hline ERRORS &
\summ{What does the system consider to be an error; how are they prevented and handled?}
Error Detection &
What errors can be detected in which feedback loops, and how? &
Human inspection in live coding vs. partial automation in static typing \\
Error Response &
How does the system respond when an error is detected? &
Does it stop, recover automatically, ignore the error or ask the user how to continue? \\
\hline ADOPTABILITY &
\summ{How does the system facilitate or obstruct adoption by both individuals and communities?}
Learnability &
What is the attitude towards the \emph{learning curve} and what is the target audience? &
HyperCard for the general public vs. FORTRAN for scientists \\
Sociability &
What are the social and economic factors that make the system the way it is? &
Funding, volunteers, code sharing, Q/A sites vs. documentation, sense of belonging \\
\hline
\end{tabular}
\caption{\label{summary-table2} Quick reference sheet for our set of Technical Dimensions.}
\end{table}

>>>>

\newcommand{\summ}[1]{\multicolumn{2}{>{\hsize=\dimexpr2\hsize+2\tabcolsep+\arrayrulewidth\relax}X}{#1} \\ \hline}
%\newcommand{\summ}[1]{\\}

\setlength\extrarowheight{2pt}

\begin{table}
\caption{Technical Dimensions cheat sheet.}
\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.75cm} X X }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline INTERACTION &
\summ{How do users manifest their ideas, evaluate the result, and generate new ideas in response?}
Feedback Loops &
How wide are the various gulfs of \emph{execution} and \emph{evaluation} and how are they related? &
Immediate Feedback (short) vs. batch mode (long) gulf of evaluation \\
Modes of Interaction &
Which sets of feedback loops only occur together? &
Setup vs. editing vs. debugging \\
Abstraction Construction &
How do we go from abstractions to concrete examples and vice versa? &
Programming by Example vs. first principles \\
\hline NOTATION &
\summ{How are the different textual / visual programming notations related?}
Notational Structure &
What notations are used to program the system and how are they related? &
Notations overlap and need sync vs. complement each other \\
Surface / Internal Notations &
What is the connection between what a user sees and what a computer program sees? &
Sequence Editing vs. Rendering, Structure Editing vs. Recovery \\
Primary / Secondary Notations &
Is one notation more important than others? &
Secondary build scripts vs. visual editor and code on equal footing in Flash \\
Expression Geography &
Do similar expressions encode similar programs? &
Concise yet error-prone vs. explicit yet verbose \\
Uniformity of Notations &
Does the notation use a small or a large number of basic concepts? &
Lisp S-expressions vs. English-like textual notations \\
\end{tabularx}
\label{tab:summary} 
\end{table}

\begin{table}
\caption{Technical Dimensions cheat sheet (continued).}
\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.5cm} X X }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline CONCEPTUAL STRUCTURE &
\summ{How is meaning constructed? How are internal and external incentives balanced?}
Conceptual Integrity vs. Openness &
Does the system present as elegantly \emph{designed} or pragmatically \emph{improvised}? &
Integrity (Everything is a X) vs. openness (compatible mixtures) \\
Composability &
What are the primitives? How can they be combined to achieve novel behaviors? &
Sequence, selection, repetition, function abstraction, recursion, logical connectives \\
Convenience &
Which wheels do users not need to reinvent? &
Small vs. expansive standard libraries \\
Commonality &
How much is \emph{common structure} explicitly marked as such? &
Common structure is redundantly flattened vs. factored out \\
\hline CUSTOMIZABILITY &
\summ{Once a program exists in the system, how can it be extended and modified?}
Staging of Customization &
Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
Externalizability / Additive Authoring &
Which portions of the system's state can be referenced and transferred to/from it? How far can the system's behavior be changed by \emph{adding} expressions? &
State is: (\emph{i}) private and hidden; (\emph{ii}) externalizable and overwriteable; (\emph{iii}) externalized and \emph{overridable} by addition \\
Self-Sustainability &
How far can the system’s behavior be changed from within? &
None (rely on external tools) vs. self-sufficient (contains everything needed) \\
\end{tabularx}
\label{tab:summary2}
\end{table}

\begin{table}
\caption{Technical Dimensions cheat sheet (continued).}
\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.5cm} X X }
\hline
Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
\hline COMPLEXITY &
\summ{How does the system structure complexity and what level of detail is required?}
Factoring of Complexity &
What programming details are hidden in reusable components and how? &
Domain-specific (more hiding) vs. general-purpose (less hiding) \\
Level of Automation &
What part of program logic does not need to be explicitly specified? &
Garbage collection (low-tech) vs. Prolog engine (hi-tech) \\
\hline ERRORS &
\summ{What does the system consider to be an \emph{error}? How are they prevented and handled?}
Error Detection &
What errors can be detected in which feedback loops, and how? &
Human inspection in live coding vs. partial automation in static typing \\
Error Response &
How does the system respond when an error is detected? &
Does it stop, recover automatically, ignore the error or ask the user how to continue? \\
\hline ADOPTABILITY &
\summ{How does the system facilitate or obstruct adoption by both individuals and communities?}
Learnability &
What is the attitude towards the \emph{learning curve} and what is the target audience? &
HyperCard for the general public vs. FORTRAN for scientists \\
Sociability &
What are the social and economic factors that make the system the way it is? &
Cathedral vs. Bazaar model \\
\end{tabularx}
\label{tab:summary3}
\end{table}
